Estimating duration in the presence of misclassification

### 2023-11-28 ATACCC-appendix-original-analysis.tex
Updates from Overleaf

This chapter describes an early analysis I did on an initial version of the ATACCC dataset.
This dataset is not publicly available.
It contains 79 individuals, all those in the study at that time with at least one positive test.
The duration distribution estimated is very similar to that in \cref{E-ATACCC} but more certain because it assumes that the random effects are drawn independently.

@@

We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
\autoref{sec:paper:hierarchy} explains the hierarchical structure which borrows information between individuals.

@@

We assume that $\log t_{i,\text{peak}}$, $C_{i,\text{peak}}$, and $\log \beta_i$ are drawn independently from univariate normal distributions (details below).
Other parameters are not hierarchical and have prior distributions placed on them in the normal way; \autoref{fig:paper:model_DAG} shows these relationships.
As in \textcite{fogartyBayesian}, a log transform is applied to the parameters which must be positive for biological plausibility.
As well as restricting these parameters to be positive, this enables the skewed distribution observed in SARS-CoV-2 with some individuals testing positive for much longer than the median.

@@

The model generally shows a good fit to the observed data (see \autoref{fig:paper:goodness_of_fit}).
For individuals with few or no observations before their peak viral load, information is borrowed from other individuals to infer when this is likely to have occurred.
Most observations, excluding those that look to be false negatives (negatives surrounded by two high viral load positives) fall within the 95\% predictive intervals.

@@

We estimate that the time from the time from when an individual has a 50\% chance of detection to peak viral load (equivalently: when $\hat{C} = C_\text{lod}$) is 2.4 (posterior median; with 95\% central CrI 1.5--3.3) days with a population standard deviation on the log scale of 0.69 (0.42--1.06).
The mean peak viral corresponds to 17.8 (17.1--18.6) Ct units, with a standard deviation of 2.2 (1.5--3.2) Ct units, after which the decrease in viral corresponds to Ct values increasing by 1.6 (1.5--1.8) Ct / day, with a standard deviation on the log scale of 0.50 (0.40--0.61).
Predictive distributions for these parameters for a new individual are shown in \autoref{fig:paper:indiv_predict}.
We estimate a 6.8 (5.0--8.8) \% probability of a false negative independent of the viral load of an individual.

@@

The time between the two points when $\hat{C} = C_\text{lod}$ for an individual can be considered a duration of infection.
This gives a mean duration estimated at 18.4 (16.5--20.9) days but the distribution is skewed, as seen by the median of 16.6 (14.8--18.5) days.
The full distribution is in \autoref{fig:paper:duration}.
\begin{figure}
    \centering
    \makebox[\textwidth][c]{\includegraphics{ATACCC/appendix/duration}}
    \caption{Proportion of people with $\hat{C} > C_\text{lod}$ a given number of days after this first occurred. Shading indicates central 95\% credible interval.}
    \label{fig:paper:duration}
\end{figure}

@@

This study presents estimates of viral load dynamics based on a general population sample with daily swabbing.
Our results show a peak Ct value around 5 Ct lower than \textcite{kisslerViral}, that is a higher viral load, although occurring at a similar time (see \autoref{tab:paper:compare}) and a longer time from first detection to peak viral load than \textcite{jonesEstimating}.
This could be because of the different populations in each study. \Textcite{kisslerViral}'s data is collected from longitudinal swabbing of individuals professionally involved in the National Basketball Association (NBA).
Of the individuals who were included in their analysis, around half were in professional athletes, likely in peak physical condition, and 90\% were male.
Our data is collected from the general population who were reported as contacts of individuals detected through surveillance open to all members of the public.
Therefore, our sample is likely to be, on average, in worse physical condition which may lead to a higher viral load (lower Ct).
Another possible difference could be the lab the tests were performed in.
Ct values are not fully comparable between different laboratories due to different testing methodologies~\autocites{dahdouhCt,hanRTPCR}.

>>>>

\todo[inline]{Check for "REF" in this appendix}

This appendix describes an early analysis I did on an initial version of the ATACCC dataset.
This dataset is not publicly available.
It contains 79 individuals, all those in the study at that time with at least one positive test.
The duration distribution estimated is very similar to that in \cref{E-ATACCC} but more certain because it assumes that the random effects are drawn independently.

@@

We assume that all observations (both of the same and different individuals) are independent conditional on the model parameters.
The parameter $\vec\theta_i$ introduces dependence between observations of the same individuals.
\Cref{sec:paper:hierarchy} explains the hierarchical structure which borrows information between individuals.

@@

We assume that $\log t_{i,\text{peak}}$, $C_{i,\text{peak}}$, and $\log \beta_i$ are drawn independently from univariate normal distributions (details below).
Other parameters are not hierarchical and have prior distributions placed on them in the normal way; \cref{fig:paper:model_DAG} shows these relationships.
As in \textcite{fogartyBayesian}, a log transform is applied to the parameters which must be positive for biological plausibility.
As well as restricting these parameters to be positive, this enables the skewed distribution observed in SARS-CoV-2 with some individuals testing positive for much longer than the median.

@@

The model generally shows a good fit to the observed data (see \cref{fig:paper:goodness_of_fit}).
For individuals with few or no observations before their peak viral load, information is borrowed from other individuals to infer when this is likely to have occurred.
Most observations, excluding those that look to be false negatives (negatives surrounded by two high viral load positives) fall within the 95\% predictive intervals.

@@

We estimate that the time from when an individual has a 50\% chance of detection to peak viral load (equivalently: when $\hat{C} = C_\text{lod}$) is 2.4 (posterior median; with 95\% central CrI 1.5--3.3) days with a population standard deviation on the log scale of 0.69 (0.42--1.06).
The mean peak viral corresponds to 17.8 (17.1--18.6) Ct units, with a standard deviation of 2.2 (1.5--3.2) Ct units, after which the decrease in viral corresponds to Ct values increasing by 1.6 (1.5--1.8) Ct / day, with a standard deviation on the log scale of 0.50 (0.40--0.61).
Predictive distributions for these parameters for a new individual are shown in \cref{fig:paper:indiv_predict}.
We estimate a 6.8 (5.0--8.8) \% probability of a false negative independent of the viral load of an individual.

@@

The time between the two points when $\hat{C} = C_\text{lod}$ for an individual can be considered a duration of infection.
This gives a mean duration estimated at 18.4 (16.5--20.9) days but the distribution is skewed, as seen by the median of 16.6 (14.8--18.5) days.
The full distribution is in \cref{fig:paper:duration}.
\begin{figure}
    \centering
    \makebox[\textwidth][c]{\includegraphics{ATACCC/appendix/duration}}
    \caption{Proportion of people with $\hat{C} > C_\text{lod}$ a given number of days after this first occurred. Shading indicates central 95\% credible interval.}
    \label{fig:paper:duration}
\end{figure}

@@

This study presents estimates of viral load dynamics based on a general population sample with daily swabbing.
Our results show a peak Ct value around 5 Ct lower than \textcite{kisslerViral}, that is a higher viral load, although occurring at a similar time (see \cref{tab:paper:compare}) and a longer time from first detection to peak viral load than \textcite{jonesEstimating}.
This could be because of the different populations in each study. \Textcite{kisslerViral}'s data is collected from longitudinal swabbing of individuals professionally involved in the National Basketball Association (NBA).
Of the individuals who were included in their analysis, around half were in professional athletes, likely in peak physical condition, and 90\% were male.
Our data is collected from the general population who were reported as contacts of individuals detected through surveillance open to all members of the public.
Therefore, our sample is likely to be, on average, in worse physical condition which may lead to a higher viral load (lower Ct).
Another possible difference could be the lab the tests were performed in.
Ct values are not fully comparable between different laboratories due to different testing methodologies~\autocites{dahdouhCt,hanRTPCR}.

### 2023-11-28 ATACCC.tex
Fix typo
Updates from Overleaf

The noisy observations are modelled by assuming each measurement of log viral load is a realisation of an iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
\todo{Paul/Dani: is this phrasing "a realisation of [a] random variable" OK?}
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored is the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with negative tests signified by $y_{i,t}=v_\text{lod}$.
Then, its likelihood is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) of false negative (bottom line).
The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.

@@

Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
Specifically, false positives are not needed to be explain the observations (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
\todo{this sentence is a bit clumsy...}
Therefore, the model for test failures can be simplified to simply have a constant probability of a false negative.
% These false negatives could represent a variety of scenarios, such as poor swabbing technique.
Formally, the model can be expressed by replacing \cref{ATACCC:eq:hakki-likelihood} with the following.
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$.}
\end{cases}
\end{align}

@@

\todo[inline]{Probably important to show estimated test sensitivity here to refer to later}

@@

While the population-level parameters converged in the modified model, both model had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
However, this was greatly improved with the simplified model, limited to a single individual, number 55.
Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estiamtes of interest.
\todo{clunky phrasing...}
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}

>>>>

The noisy observations are modelled by assuming each measurement of log viral load is a realisation of an iid Normally-distributed random variable error with mean 0 and standard deviation $\sigma_v$.
\todo{Paul/Dani: is this phrasing "a realisation of [a] random variable" OK?}
The measurements are left-censored at the limit of detection, $\vlod$, meaning that any observation which would give a log viral load of less than $\vlod$ is instead observed as negative.
False positives and negatives are modelled by allowing a test failure with probability $\rho$.
Test failures are modelled as producing a log viral load reading drawn from $\N(x_0, \sigma_0^2)$ where $x_0$ and $\sigma_0^2$ are parameters to be estimated.
These readings are censored in the same way as true readings.
A censored test failure represents a false negative, otherwise the test failure represents a false positive.
\textcite{hakkiOnset} uses a prior for $x_0$ centred on $\vlod$, which encodes a prior belief that false negatives and false positives occur with equal probability.

Let the observed log viral load for individual $i$ at time $t$ be $y_{i,t}$, with $y_{i,t}=0$ if the test is negative.
Then, its likelihood is:
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    \rho f_N(y_{i,t} \mid x_0, \sigma_0^2) + (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho F_N(v_\text{lod} \mid x_0, \sigma_0^2) + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$} \label{ATACCC:eq:hakki-likelihood}
\end{cases}
\end{align}
where $f_N(y \mid \mu, \sigma^2)$ and $F_N(y \mid \mu, \sigma^2)$ are the pdf and cdf at $y$ of a Normal distribution with mean $\mu$ and variance $\sigma^2$ respectively.
The first term on each line is the likelihood contribution from test failures, either a false positive (top line) or false negative (bottom line).
The second term on the first line is the likelihood of the observed log viral load from non-failed test results.
The second term on the bottom line is the likelihood of the negative (censored) tests from non-failed test results.

@@

Based on the fit to the data achieved when implementing an earlier version of the model (see \cref{ATACCC:sec:original-analysis}), I hypothesised that the cause of this was overcomplicating the model for test failures.
Specifically, false positives are not needed to be explain the observations (the results in \cref{fig:paper:duration,ATACCC:fig:goodness-of-fits} do not allow false negatives but fit the data well).
\todo{this sentence is a bit clumsy...}
Therefore, the model for test failures can be simplified to have a constant probability of a false negative.
% These false negatives could represent a variety of scenarios, such as poor swabbing technique.
Formally, the model can be expressed by replacing \cref{ATACCC:eq:hakki-likelihood} with the following.
\begin{align}
p(y_{i,t} \mid \cdot) &= \begin{cases}
    (1 - \rho) f_N(y_{i,t} \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} > v_\text{lod}$}\\
    \rho + (1 - \rho) F_N(\vlod \mid \log v_i(t), \sigma_v^2) & \text{if $y_{i,t} = v_\text{lod}$.}
\end{cases}
\end{align}

@@

@@

While the population-level parameters converged in the modified model, both models had convergence issues (Rhat > 1.4 and ESS < 20) for some individual-level parameters.
However, this was greatly improved with the simplified model, limited to a single individual, number 55.
Individual 55 has an extremely noisy set of observations.
This means that two viral load trajectories fit the data well (shown in \cref{ATACCC:fig:individual-55}).
The first, found by two chains, increases very quickly, peaks 4.0 days after the first test, then declines slowly.
The second, found by the other six chains, is more symmetrical, peaking 7.2 days after the first test.
The individual trajectories are not of interest for this analysis, and the population-level parameters agree between the two sets of chains, therefore this should not affect any estimates of interest.
\todo{clunky phrasing...}
\begin{figure}
  \centering \includegraphics{ATACCC/fit_individual_55}
  \caption[Goodness-of-fit for individual 55]{Fit to individual 55's data coloured by MCMC chain. Two chains favour the quickly increasing viral load trajectory, peaking at day 4.0, the other six favour the slower trajectory, peaking at day 7.2. \label{ATACCC:fig:individual-55}}
\end{figure}
